Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-02 paper.tex
Briefly define HMM consistency when we first say it

In the wide ARG (sampled to mid-2021), all these sequences are descendants of
recombination nodes, except MILK-103C712, which was removed during
preprocessing (Table 2); additionally, all are HMM-consistent, except
QEUH-1067DEF. Overall, we find excellent agreement between Jackson et al.’s
motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
and (2) the breakpoint intervals (Table 2). In all cases, the two methods
proposed identical or very closely related parent lineages. The general mosaic
genome structure is concordant in all the cases, even CAMC-CB7AB3 where there
are two switches in the HMM copying paths (indicative of two breakpoints). The
breakpoint intervals estimated using the two methods overlap substantially in
all the cases (Table 2). Within each of the groups A to D, the constituent
sequences have identical breakpoint intervals, because they trace back to a
single recombination node in the wide ARG. This result is consistent with the
hypothesis that the sequences of the groups A to D emerged from four separate
recombination events~\cite{Jackson2021-ik}.

@@

As a check on the reliability of these identified recombination events,
Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are
"HMM-consistent"; that is, where the mirrored HMM runs (i.e., 5’ to 3’ end
versus 3’ to 5’ end) identify the same set of originating mutations and the
same number of breakpoints, and the parents in the mirrored runs are
sequence-identical. This reduces both the number of recombination nodes /
breakpoints from 1522 / B to C / D in the wide ARG and from 763 / 806 to 459 in
the long ARG (Supplementary Tables 1 and 2).

>>>>

In the wide ARG (sampled to mid-2021), all these sequences are descendants of
recombination nodes, except MILK-103C712, which was removed during
preprocessing (Table 2). Additionally, all except QEUH-1067DEF are HMM-consistent:
that is, the mirrored HMM runs (i.e. forward from 5’ to 3’ end versus backwards from 3’ to 5’ end)
identify the same set of originating mutations, have the same number of breakpoints, and have
equivalent (sequence-identical) parents from forward and backward runs.
Overall, we find excellent agreement between Jackson et al.’s
motif-based method and sc2ts in terms of (1) the Pango lineage of the parents
and (2) the breakpoint intervals (Table 2). In all cases, the two methods
proposed identical or very closely related parent lineages. The general mosaic
genome structure is concordant in all the cases, even CAMC-CB7AB3 where there
are two switches in the HMM copying paths (indicative of two breakpoints). The
breakpoint intervals estimated using the two methods overlap substantially in
all the cases (Table 2). Within each of the groups A to D, the constituent
sequences have identical breakpoint intervals, because they trace back to a
single recombination node in the wide ARG. This result is consistent with the
hypothesis that the sequences of the groups A to D emerged from four separate
recombination events~\cite{Jackson2021-ik}.

@@

As a check on the reliability of these identified recombination events,
Figure~\ref{fig:recomb_mrcas}B  plots only those breakpoints which are
HMM-consistent. This reduces both the number of recombination nodes /
breakpoints from 1522 / B to C / D in the wide ARG and from 763 / 806 to 459 in
the long ARG (Supplementary Tables 1 and 2).
